Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-06 paper/paper.tex
rough draft of syntax section

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathsf{string}}
\newcommand{\charg}{\mathsf{char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}

@@

@@

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\el}{\mathsf{el}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
\newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
\newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

\newcommand{\ctxwff}[1]{#1 \,\, \mathsf{ok}}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{type}}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \,\, \mathsf{ok}}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \,\, \mathsf{linear}}

@@

\max{was this paragraph supposed to be deleted?}
Next we present the full details of the syntax and typing of \theoryname.
\theoryname has two kinds of types, linear types $A$ which represent formal
grammars, and non-linear types $X$, which act like ordinary types or sets. Both
linear and non-linear types are checked for well-formedness in a context
$\Gamma$ containing \emph{only} non-linear variables $\Gamma = x:X,\ldots$.

\begin{figure}
  \[
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
                                                  \mathsf{Int} \pipe
                                                  \mathsf{Empty} \pipe
                                                  \mathsf{Unit} \pipe
                                                  \mathsf{SPF}\,X\pipe U \pipe L
                                                  \pipe \ltonl A \pipe \sum\limits_{x:X} Y
                                                  \pipe \prod\limits_{x:X} Y\pipe M =_A
                                                  N \\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
  \end{array}
  \]
  \caption{Overview of Types}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma ; \Delta \vdash e : A}

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    {\Gamma \vdash M : \ltonl A}
    {\Gamma ; \cdot \vdash M : A}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}
  \caption{Linear typing}
  \label{fig:linsyntax}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash \Set_0 : \Set_1}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
    {}{}
  \end{mathpar}
  \caption{Universe Types}
\end{figure}

\begin{figure}
  \textrm{TODO: var, konstant, bigoplus, bigamp and tensor}\\
  \textrm{TODO: behavior of map}
  \begin{mathpar}
    \inferrule
    {\Gamma \vdash M : X}
    {\Gamma \vdash \mathsf{Var}\,M : \SPF\,X}

    \inferrule
    {\Gamma \vdash A : L}
    {\Gamma \vdash \mathsf{K}\,A : \SPF\,X}

    \inferrule
    {\Gamma,y:Y \vdash A : \SPF\,X}
    {\Gamma \vdash \bigoplus_{y:Y}\,A : \SPF\,X}

    \inferrule
    {\Gamma,y:Y \vdash A : \SPF\,X}
    {\Gamma \vdash \bigamp_{y:Y}\,A : \SPF\,X}

    \inferrule
    {\Gamma \vdash A : \SPF\,X\and \Gamma \vdash B : \SPF\,X}
    {\Gamma \vdash A \otimes B : \SPF\,X}

    \inferrule
    {A : \SPF\,X}
    {\el(A) : (X \to L) \to L}

    \el(\Var\,M) B = B\,M\and
    \el(\mathsf{K}\,A) B = A\and
    \el(\bigoplus_{y:Y} A) B = \bigoplus_{y:Y}\el(A)B\and
    \el(\bigamp_{y:Y} A) B = \bigamp_{y:Y}\el(A)B\and
    \el(A \otimes A') B = \el(A)B \otimes \el(A')B

  \inferrule
  {A : \textrm{SPF} X\and
  \Gamma,x:X;b: B x \vdash f : C x\and
  \Delta \vdash e : \el(A)(B)
  }
  {\Gamma;\Delta \vdash \mathsf{map}\,A\,(x.b.f)\,e : \textrm{el}(A)\, C}

  \map(\Var\,M)\,(x.b.f)\,e = f[M/x][e/b]\and
  \map(\K\,A)\,(x.b.f)\,e = e\and
  \map(\bigoplus\limits_{y:Y} A)\,(x.b.f)\,e =
  \letin{\oplusinj y a}{e}{\oplusinj y {\map(A)\,(x.b.f)\,e}}\\
  \map(\bigwith\limits_{y:Y} A)\,(x.b.f)\,e =
  \withlamb{y}{\map(A)\,(x.b.f)\,(\withprj a e)}\\
  \map(A \otimes A')\,(x.b.f)\,e =
  \letin{(a,a')}{e}{(\map(A)\,(x.b.f)\,a,\map(A')\,(x.b.f)\,a')}
  \end{mathpar}
  \caption{Strictly Positive Functors}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=IndIntro]
    {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
     \Gamma \vdash M : X\and
     \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
    {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}

    \inferrule*[right=IndElim]
    {\Gamma;\Delta \vdash e : \mu A M\and
     \Gamma,x:X;b:\el(A\,x)\,B \vdash f : B\,x
    }
    {\Gamma; \Delta \vdash \mathsf{fold}(A)(x.b.f)(M, e) : B\,M}

    \inferrule*[right=Ind$\beta$]
    {}
    {\fold(A)(x.b.f)(M, \roll(e)) =
      f[M/x][\map(A\,M)\,(y.a.\fold(A)(x.b.f)(y,a))\,e/b]}

    \inferrule*[right=Ind$\eta$]
    {\Gamma,x:X; a:\mu\,A\,x \vdash e' : B\,x
      \and \Gamma;\Delta \vdash e : \mu A M\\\\
      \Gamma,x:X; a':\el(A x)(\mu A) \vdash
      e'[x/x][\roll(a')/a] = f[x/x][\map(A\,x)\,(y.a.e'[y/x][a/a])\,a/b] : B\,x
    }
    {\Gamma; \Delta \vdash \fold(A)(x.b.f)(M,e) = e'[M/x][e/a] : B\,M}
  \end{mathpar}
  \caption{Indexed Inductive Linear Types}
\end{figure}

>>>>

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathsf{string}}
\newcommand{\charg}{\mathsf{char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\newcommand{\quoteTy}[1]{\lceil{#1}\rceil}
\newcommand{\unquoteTy}[1]{\lfloor{#1}\rfloor}

@@

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\el}{\mathsf{el}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\withlamb}[2]{{\lambda}^{{\&}} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \, #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\textstyle\prod (#1 : #2). #3}
\newcommand{\SigTy}[3]{\textstyle\sum (#1 : #2). #3}
\newcommand{\LinPiTy}[3]{\textstyle\bigamp (#1 : #2). #3}
\newcommand{\LinPiTyLimit}[3]{\bigwith\limits_{#1 : #2} #3}
\newcommand{\LinSigTy}[3]{\textstyle\bigoplus (#1 : #2). #3}
\newcommand{\LinSigTyLimit}[3]{\bigoplus\limits_{#1 : #2} #3}
%% \newcommand{\DepWith}[2]{{\textstyle\bigamp}\limits_{#1}{#2}
%% \newcommand{\DepPlus}[2]{{\textstyle\bigoplus}\limits_{#1}{#2}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

\newcommand{\ctxwff}[1]{#1 \isCtx}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \isTy}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \isLinCtx}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}

@@

%% OUTLINE
%%
%% 1. Overview of syntax: what are the judgments
%% 2. Cover the linear types except for inductives
%% 3. Cover non-linear types, mainly focus on the universes
%% 4. Cover indexed inductive linear types
%% 5. Cover the "axioms" we add

\max{was this paragraph supposed to be deleted?}  Next we present the
full details of the syntax and typing of \theoryabbv. The syntax of
\theoryabbv is based on the dependent linear-non-linear (\lnld)
calculus from \cite{dlnl}. As in \lnld, \theoryabbv includes both
non-linear dependent types, as well as linear types, which are allowed
to depend on the non-linear types, but not on other linear types.
%
The main point of departure from \lnld's design in that the linear
typing is \emph{non-commutative}, i.e., that exchange is not an
admissible structural rule. Furthermore, we add a general-purpose
indexed inductive linear type connective, as well as an
\emph{equalizer} type, which we will show allows us to perform
inductive proofs of equalities between linear terms.
%
Finally, we add base types and axioms to the system specifically to
model formal grammars and parsing.

The formation rules for the judgments of \theoryabbv are shown in
\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
stand for non-linear types, $M,N$ stand for non-linear terms, these
act as in an ordinary dependent type theory. $\Delta$ stands for
linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
terms. These contexts, types and terms are allowed to depend on an
ambient non-linear context $\Gamma$, but note that linear types $A$
cannot depend on any \emph{linear} variables in $\Delta$. We include
definitional equality judgments for both kinds of type and term
judgments as well. Additionally, we have a judgment $\Gamma\vdash X
\isSmall$ which is used in the definition of universe types.

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\ctxwff \Gamma}

    \inferrule{\ctxwff \Gamma}{\ctxwffjdg \Gamma X}

    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}

    \inferrule{\ctxwffjdg \Gamma X \and \ctxwffjdg \Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}

    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash M : X}

    \inferrule{\Gamma \vdash M : X \and \Gamma \vdash N : X}{\Gamma \vdash M \equiv N : X}

    \inferrule{\ctxwff \Gamma}{\linctxwff \Gamma \Delta}

    \inferrule{\ctxwff \Gamma}{\Gamma \vdash A \isLinTy}

    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \equiv B}

    \inferrule{\Gamma \vdash \Delta \isLinCtx \and \Gamma \vdash A \isLinTy}{\Gamma;\Delta\vdash e : A}

    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
  \end{mathpar}
  \caption{Formation Rules}
  \label{fig:formation}
\end{figure}

\subsection{Non-linear Typing}

The non-linear types of \theoryabbv include the standard dependent
types for $\Pi,\Sigma$, extensional equality, natural numbers,
booleans, unit and empty types. We include these rules in the
appendix. We present the other non-linear type constructions in
\Cref{fig:non-linear-stuff}. First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
``universes ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment. These universe types are needed so that we
can define types by recursion on natural numbers. Next, we include a
non-linear type $\ltonl A$ where $A$ is a linear type. The intuition
for this type is that its elements are the linear terms that are
``resource free'': its introduction rule says we can construct an
$\ltonl A$ when we have a linear term of type $A$ with no free linear
variables. This type is used extensively in our examples, playing a
similar role to the $!$ modality of ordinary linear logic.

%% \begin{figure}
%%   \[
%%   \begin{array}{rrcl}
%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
%%                                                   \mathsf{Bool} \pipe
%%                                                   \mathsf{Empty} \pipe
%%                                                   \mathsf{Unit} \pipe
%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
%%                                                   N
%%   \end{array}
%%   \]
%%   \caption{Non-linear Types}
%% \end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash U \isTy}\and
    \inferrule
    {\Gamma \vdash M : U}
    {\Gamma \vdash \unquoteTy M \isTy}\and
    \inferrule
    {\Gamma \vdash X \isTy}
    {\Gamma \vdash \quoteTy X : U}\and
    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}

    \inferrule{}{\Gamma \vdash L \isTy}\and
    \inferrule
    {\Gamma \vdash M : L}
    {\Gamma \vdash \unquoteTy M \isLinTy}\and
    \inferrule
    {\Gamma \vdash A \isLinTy}
    {\Gamma \vdash \quoteTy A : L}\and
    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}

    \inferrule
    {\Gamma \vdash A \isLinTy}
    {\Gamma \vdash \ltonl { A } \isTy}
    \and
    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
  \end{mathpar}
  \caption{Non-linear types (selection)}
  \label{fig:non-linear-types}
\end{figure}

\subsection{Linear Typing}

We give an overview of the linear types in \Cref{fig:linear-types} and
terms in \Cref{fig:linear-terms}. The equational theory for these
types is straightforward and included in the appendix. First, the
linear variable rule says that a linear variable can be used if it is
the \emph{only} variable in the context.

First, we cover the ``multiplicative'' connectives of non-commutative
linear logic. The linear unit and tensor product are standard for a
non-commutative linear logic: when we construct a linear unit we
cannot use any variables and when we construct a tensor product, the
two sides must use disjoint variables, and the variables the left side
of the product uses must be to the left in the context of the
variables used by the right side of the tensor product. Because we are
non-commutative, there are two function types: $A \lto B$ and $B \tol
A$. The difference between these is that the introduction rule for $A
\lto B$ adds a variable to the right side of the context, whereas the
introduction rule for $B \tol A$ adds a variable to the left side of
the context. In our experience, because by convention parsing
algorithms parse from left-to-right, we rarely need to use the $B \tol
A$ connective. One of the main uses of linear types is that, in
combination with the $\uparrow$ connective, we can abstract over
linear terms, in that the non-linear type $\ltonl{A \lto B}$ is a
non-linear type of linear terms with one linear input $A$ and output
$B$.

Next, we have the ``additive'' connectives. For these, we use the
non-linear types to define \emph{indexed} versions of the additive
disjunction $\oplus$ and additive conjunction $\&$ of linear
logic. These are like a linear version of the $\Sigma$ and $\Pi$
connectives of ordinary dependent type theory. The indexed $\&$ is
defined by a $\lambda$ and eliminated using projection. The indexed
$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
introduction rule $\sigma$, but its elimination rule is given by
\emph{pattern matching} rather than projections. Note that we can
define the more typical nullary and binary versions of these
connectives by using indexing over the empty and boolean type
respectively. We will freely use $\bot$ to refer to this empty
discunction and $\top$ to refer to the empty conjunction. Lastly, we
include a type $\equalizer {a}{f}{g}$ that we call the
\emph{equalizer} of $f$ and $g$. We think of this type as the
``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
these were non-linear types, such a type could be constructed using
the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
can't be used because it uses a \emph{dependent} equality type. We
note that the equalizer type is not used in defining any of our
parsers or formal grammars, but instead will be used shortly to derive
an induction principle for proving equalities of functions whose
domain is an inductive type.

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash I \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A}\and
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
  \end{mathpar}
  \caption{Linear Types (selection)}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A e x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {e}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}
  \caption{Linear terms (selection)}
  \label{fig:linsyntax}
\end{figure}

\subsection{Indexed Inductive Linear Types}

Next, we introduce one of the most complex and important type linear
type constructor of our development, \emph{indexed inductive linear
types} $\mu F$. To encode these, we first add a non-linear type
$\SPF\,X$ of \emph{strictly-positive functorial} linear type
expressions indexed by a non-linear type $X$. We give the rules for
strictly-positive functors in \Cref{fig:spf}. First, we add operations
for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
products and additive conjunction and disjunction. As an example,
\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
interprets our syntactic codes as an actual type constructor, and
define it for each of the operations given above. Lastly, we give a
\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
  find a citation for this kind of approach. Container papers work
  this way for instance. I think there is a linear container paper
  that uses this appraoch}

With this notion of strictly-positive functor so defined, in
\Cref{fig:iilt} we next add a linear type constructor $\mu F$, which
defines a family of mutually-recursive inductive types for the family
of type expressions $F$. The introduction rule for this is
$\mathsf{roll}$, which constructs an element of $\mu F\,x$ from its
one-level unfolding. The elimination principle is defined by a
$\mathsf{fold}$ operation. The equations ensure that this makes the
family an \emph{initial algebra} for the type expressions. That is the
$\beta$ rule says that a $\mathsf{fold}$ applied to a $\mathsf{roll}$
is equivalent to $\mathsf{map}$ping the $\mathsf{fold}$ over all the
sub-expressions, which means that $\mathsf{fold}$ is an algebra
homomorphism from $\mathsf{roll}$ to $\mathsf{f}$. Then the $\eta$
rule says that $\mathsf{fold}$ is the \emph{unique} such homomorphism.

This definition as an initial algebra is well-understood semantically
but the $\eta$ principle in particular is somewhat cumbersome to use
directly in proofs. In dependent type theory, we would have a
dependent \emph{elimination} principle, which justifies
\emph{inductive proofs}. However since linear types do not support
dependency on linear types, we cannot directly adapt this
approach. However, if we are trying to prove that two morphisms out of
a mutually recursive type are equal, we can use the \emph{equalizer}
type to prove their equality by induction.\max{TODO: show how this is
  done.}

\begin{figure}
  \begin{align*}
    \mathsf{Var} &: X \to \SPF\,X\\
    \mathsf{K} &: L \to \SPF\,X\\
    \mathsf{\bigoplus} &: (Y \to \SPF\,X) \to \SPF\,X\\
    \mathsf{\bigamp} &: (Y \to \SPF\,X) \to \SPF\,X\\
    \mathsf{\otimes} &: \SPF\,X \to \SPF\,X \to \SPF\,X\\
    \el &: \prod_{X:U}\SPF\,X \to (X \to L) \to L\\
    \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}(\prod_{x:X}\uparrow(A\,x\lto B\,x)) \to \uparrow(\el(F)(A) \lto \el(F)(B))\\\\
    \el(\Var\,M) B &= B\,M\\
    \el(\mathsf{K}\,A) B &= A\\
    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
    \map(\Var\,M)\,f &= f\,M\\
    \map(\K\,A)\,f &= \lambda a. a\\
    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
  \end{align*}
  \begin{mathpar}
  \end{mathpar}
  \caption{Strictly Positive Functors}
  \label{fig:spf}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \mathsf{roll} : \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}

    %% \inferrule*[right=IndIntro]
    %% {\Gamma \vdash A : X \to \textrm{SPF}\,X\and
    %%  \Gamma \vdash M : X\and
    %%  \Gamma; \Delta \vdash e : \textrm{el}(A\,M) (\mu\,A)}
    %% {\Gamma; \Delta \vdash \mathsf{roll}\, e : \mu\,A\, M}

    \mathsf{fold} : \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
    (\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)})
    \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}

    \inferrule*[right=Ind$\beta$]
    {}
    {\fold\,F\,f\,x\,(\roll e) \equiv f\,x\,(\map(F\,x)\,(\fold\,F\,f))}
